PCA

### 2019-12-20 analysis/02_PCA.Rmd
new pca and tsne plots

@@

```{r read}
pseudocounts <- read_csv("../results/01_pseudo.counts.csv")
head(pseudocounts[1:3])
```

## Characterization only

```{r characterization}
colData <- read.csv("../metadata/00_colData_characterization.csv", header = T, row.names = 1)
colData$treatment <- factor(colData$treatment, levels = charlevels)
colData <- colData %>% mutate(tissue = fct_recode(tissue, "gonads" = "gonad"))
colData$tissue <- factor(colData$tissue, levels = tissuelevels)

head(colData[1:3])

geneinfo <- read.csv("../metadata/00_geneinfo.csv", row.names = 1)
head(geneinfo)

subsetcounts <- function(colData, countData){

  # save counts that match colData
  savecols <- as.character(colData$V1) 
  savecols <- as.vector(savecols) 
  countData <- countData %>% 
    dplyr::select(X1,savecols) %>% 
    dplyr::rename(entrezid = X1)

  countData <- left_join(geneinfo, countData)
  countData <- countData %>% distinct(Name, .keep_all = TRUE) %>%
    select(-row.names)  %>% select(-geneid) %>% select(-entrezid)

  row.names(countData) <- countData$Name
  countData$Name <- NULL
  countData <- as.data.frame(t(countData))

  return(countData)
}

countData <- subsetcounts(colData, pseudocounts)
head(countData[1:3])

# confirm
#row.names(countData) == row.names(colData)

euclidist <- dist(countData) # euclidean distances between the rows

# make pca dataframes
mypca <- prcomp(countData)

# percent conribution
eig.val <- get_eigenvalue(mypca)
head(eig.val)

@@

a <- ggplot(mypcadf, aes(x = PC1, y = PC2, 
                         color = colData$treatment, 
                         shape = colData$tissue)) +
  geom_point()   +
  mytheme() +
  theme(legend.title = element_blank(), 
        axis.text = element_blank()) +
  scale_alpha_manual(values = c(0.5,1)) +
  labs(x = "PC1: 35.9% of variance", y = "PC2: 32.5% of variance")

b <-  fviz_pca_var(mypca,  labelsize = 3 , axes.linetype = "blank", 
                   repel = TRUE ,
                  select.var= list(contrib = 6)) + 
      mytheme() + 
      labs(x = " ", y = " ", subtitle = NULL, title =  NULL) +
    theme( axis.text = element_blank())

plot_grid(a,b, rel_widths = c(2,1))

fviz_screeplot(mypca, addlabels = TRUE, ylim = c(0, 50))
fviz_contrib(mypca, choice = "var", axes = 1, top = 5) 
fviz_contrib(mypca, choice = "var", axes = 2, top = 5)

tsne_model <- Rtsne(euclidist, check_duplicates=FALSE, pca=TRUE, perplexity=5, theta=0.5, dims=2)
tsne_df = as.data.frame(tsne_model$Y)

c  <- ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = sex, color = tissue)) +
  geom_point(size = 2) +
  mytheme() +
  scale_color_manual(values = c("#d95f02","#1b9e77", "#7570b3")) +
  labs(x = "tSNE 1", y = "tSNE 2")

d <- ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = tissue, color = treatment)) +
  geom_point(size = 2) +
  mytheme() +
  labs(x = "tSNE 1", y = "tSNE 2") +
  scale_alpha_manual(values = c(0.5,1))

c
d
```

## Manipulations only

```{r manipulation}
head(pseudocounts[1:3])

colData <- read.csv("../metadata/00_colData_manipluation.csv", header = T, row.names = 1)

colData$treatment <- factor(colData$treatment, levels = maniplevels1)
colData <- colData %>% mutate(tissue = fct_recode(tissue, "gonads" = "gonad"))
colData$tissue <- factor(colData$tissue, levels = tissuelevels)

head(colData[1:3])

countData <- subsetcounts(colData, pseudocounts)

# confirm
row.names(colData) <- colData$V1
#row.names(countData) == row.names(colData)

euclidist <- dist(countData) # euclidean distances between the rows

# make pca dataframes
mypca <- prcomp(countData)

# percent conribution
eig.val <- get_eigenvalue(mypca)
head(eig.val)

mypcadf <- data.frame(PC1 = mypca$x[, 1], PC2 = mypca$x[, 2], PC3 = mypca$x[, 3], 
                  PC4 = mypca$x[, 4],PC5 = mypca$x[, 5],PC6 = mypca$x[, 6],
                  ID = row.names(countData))
mypcadf$V1 <- row.names(mypcadf)
mypcadf <- left_join(colData, mypcadf)
mypcadf <- mypcadf %>% dplyr::select(bird,sex,tissue,treatment,PC1:PC6)
mypcadf$treatment <- factor(mypcadf$treatment, levels = maniplevels1)
head(mypcadf)

a <- ggplot(mypcadf, aes(x = PC1, y = PC2, 
                         color = colData$treatment, 
                         shape = colData$tissue)) +
  geom_point(size = 2)  + 
  mytheme() +
  theme(legend.title = element_blank(), 
        axis.text = element_blank()) +
  scale_alpha_manual(values = c(0.5,1)) +
  scale_color_manual(values = colorscharmaip2) +
  labs(x = "PC1: 63.6% of variance", y = "PC2: 30.1% of variance")

b <-  fviz_pca_var(mypca,  labelsize = 3 , axes.linetype = "blank", 
                   repel = TRUE ,
                  select.var= list(contrib = 6)) + 
      mytheme() + 
      labs(x = " ", y = " ", subtitle = NULL, title =  NULL) +
  theme( axis.text = element_blank())

plot_grid(a,b, rel_widths = c(2,1))

fviz_screeplot(mypca, addlabels = TRUE, ylim = c(0, 50))
fviz_contrib(mypca, choice = "var", axes = 1, top = 5) 
fviz_contrib(mypca, choice = "var", axes = 2, top = 5)

tsne_model <- Rtsne(euclidist, check_duplicates=FALSE, pca=TRUE, perplexity=5, theta=0.5, dims=2)
tsne_df = as.data.frame(tsne_model$Y)

c  <- ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = sex, color = tissue)) +
  geom_point(size = 2) +
  mytheme() +
  scale_color_manual(values = c("#d95f02","#1b9e77", "#7570b3")) +
  labs(x = "tSNE 1", y = "tSNE 2")

d <- ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = tissue, color = treatment)) +
  geom_point(size = 2) +
  mytheme() +
  labs(x = "tSNE 1", y = "tSNE 2") +
  scale_color_manual(values = colorscharmaip2) +
  scale_alpha_manual(values = c(0.5,1))

c
d
```

```{r charmanip}
# prep col data for all samples
colData <- read.csv("../metadata/00_samples.csv", header = T, row.names = 1)
colData$treatment <- factor(colData$treatment, levels = alllevels3)
colData <- colData %>% mutate(tissue = fct_recode(tissue, "gonads" = "gonad"))
colData$tissue <- factor(colData$tissue, levels = tissuelevels)
row.names(colData) <- colData$V1

# prep count data for all samples
countData <- as.data.frame(pseudocounts)
countData$entrezid <- countData$X1
countData <- left_join(geneinfo, countData)
countData <- countData %>% distinct(Name, .keep_all = TRUE) %>%
    select(-row.names, -geneid, -entrezid, -X1) 
countData <- as.data.frame(countData)
row.names(countData) <- countData$Name
countData$Name <- NULL
countData$entrezid <- NULL
countData <- as.data.frame(t(countData))

# check ready for analysis
#row.names(countData) == row.names(colData)

# pca
mypca <- prcomp(countData)
eig.val <- get_eigenvalue(mypca)
head(eig.val)

a <- ggplot(mypcadf, aes(x = PC1, y = PC2, 
                         color = colData$treatment, 
                         #shape = colData$tissue,
                         #size = colData$sex,
                        shape = colData$study)) +
  #geom_hline(yintercept=0, color = "grey") +
  #geom_vline(xintercept=0, color = "grey") +
  geom_point()  +
  mytheme() +
  theme(legend.title = element_blank(),
        axis.text = element_blank()) +
  #scale_alpha_manual(values = c(0.5,1)) +
  labs(x = "PC1: 44.6% of variance", y = "PC2: 29.8% of variance") +
  scale_color_manual(values = colorscharmaip)

b <- fviz_pca_var(mypca,  labelsize = 4 , axes.linetype = "blank", 
                   repel = TRUE ,
                  select.var= list(contrib = 6)) + 
      mytheme() + 
      labs(x = " ", y = " ", subtitle = NULL, title =  NULL) +
    #geom_hline(yintercept=0, color = "grey") +
  #geom_vline(xintercept=0, color = "grey") +
    theme( axis.text = element_blank())

fviz_screeplot(mypca, addlabels = TRUE, ylim = c(0, 50))
fviz_contrib(mypca, choice = "var", axes = 1, top = 5) 
fviz_contrib(mypca, choice = "var", axes = 2, top = 5)

plot_grid(a,b,rel_widths = c(2,1))

# prep for tsne
euclidist <- dist(countData) # euclidean distances between the rows

@@

ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = sex, color = tissue)) +
  geom_point(size = 2) +
  mytheme() +
  scale_color_manual(values = colorstissue) +
  labs(x = "tSNE 1", y = "tSNE 2")

ggplot(tsne_df_cols, aes(x = V1, y = V2, shape = study, color = treatment)) +
  geom_point(size = 2) +
  mytheme() +
  labs(x = "tSNE 1", y = "tSNE 2") +
  scale_alpha_manual(values = c(0.5,1)) +
  scale_color_manual(values = colorscharmaip2)
```

>>>>

# note. this file is too big for storage in github :(

@@

```{r read}
pseudocounts <- read_csv("../results/01_pseudo.counts.csv")
head(pseudocounts[1:3])
pseudocounts <- as.data.frame(pseudocounts)
row.names(pseudocounts) <- pseudocounts$X1
pseudocounts$X1 <- NULL
# prep count data for all samples
countData <- as.data.frame(t(pseudocounts))
head(countData[1:3])
```

```{r colData}
# prep col data for all samples
colData <- read.csv("../metadata/00_samples.csv", header = T, row.names = 1)
colData$treatment <- factor(colData$treatment, levels = alllevels)
colData <- colData %>% mutate(tissue = fct_recode(tissue, "gonads" = "gonad"))
colData$tissue <- factor(colData$tissue, levels = tissuelevels)
row.names(colData) <- colData$V1

# check ready for analysis
#row.names(countData) == row.names(colData)
```

```{r pca}
# pca
mypca <- prcomp(countData)
eig.val <- get_eigenvalue(mypca)
head(eig.val)

@@

plotcolorfulpcs <- function(whichfactor, whichcolors){
  p <- ggplot(mypcadf, aes(x = PC1, y = PC2, color = whichfactor)) +
    geom_point(size = 1)  +
    theme_B3() +
    theme(legend.title = element_blank(),
         axis.text = element_blank(),
         legend.position = "none") +
    labs(x = "PC1", y = "PC2") +
    scale_color_manual(values = whichcolors) 
  print(p)
}

a <- plotcolorfulpcs(mypcadf$tissue, colorstissue)  
b <- plotcolorfulpcs(mypcadf$sex, sexcolors) 
c <- plotcolorfulpcs(mypcadf$treatment, colorscharmaip)

d <- fviz_screeplot(mypca, addlabels = TRUE, ylim = c(0, 50),  ncp = 5) + labs(title = NULL, y = "Variance") + theme_B3() 
e <- fviz_contrib(mypca, choice = "var", axes = 1, top = 5)  + labs(title = NULL, subtitle = "PC1", x = NULL) + 
          theme_B3() + theme(axis.text.x = element_text(angle = 55, hjust = 1))
f <- fviz_contrib(mypca, choice = "var", axes = 2, top = 5)  + labs(title = NULL, subtitle = "PC2", y = NULL, x = NULL) + 
          theme_B3()  + theme(axis.text.x = element_text(angle = 55, hjust = 1))
g <- fviz_pca_var(mypca,  labelsize = 3 , axes.linetype = "blank", 
                   repel = TRUE ,
                  select.var= list(contrib = 6)) + 
      theme_B3() + 
      labs(x = "PC1", y = "PC2", title =  NULL) +
      theme( axis.text = element_blank())

abc <- plot_grid(a,b, c, nrow =  1,  labels = c("d", "e", "f"), label_size = 12)
defg <- plot_grid(d,e,f,g, nrow = 1, rel_widths = c(1, 0.55, 0.45, 1), labels = c("h", "i", "j", "k" ), label_size = 12)
dg <- plot_grid(d,g, NULL, nrow = 1, rel_widths = c(1, 1, 1), labels = c("h", "i", NULL ), label_size = 12)
plot_grid(abc, defg, nrow = 2)
```

```{r tSNE}
# prep for tsne
euclidist <- dist(countData) # euclidean distances between the rows

@@

plotcolorfultsnes <- function(whichfactor, whichcolors){
  p <- ggplot(tsne_df_cols, aes(x = V1, y = V2, color = whichfactor)) +
    geom_point(size = 1) +
    theme_B3() +
    labs(x = "tSNE 1", y = "tSNE 2") +
    scale_color_manual(values = whichcolors) +
    theme(legend.position = "none",
          axis.text = element_blank())
  print(p)
}

h <- plotcolorfultsnes(tsne_df_cols$tissue, colorstissue)  
i <- plotcolorfultsnes(tsne_df_cols$sex, sexcolors)   
j <- plotcolorfultsnes(tsne_df_cols$treatment, colorscharmaip)

hij <- plot_grid(h,i,j, nrow = 1, labels = c("a", "b", "c"), label_size = 12)
hij
```
```{r PCA-tSNE}
plot_grid(hij, abc, dg, nrow = 3)
```
